---
title: "DEGs analysis based on contrinuous estradiol level in PBMC dataset"
output: html_notebook
---

```{r}
rm(list = ls())  # Removes all variables and functions from global environment
getwd()
```

```{r}
setwd("C:/Users/ngduy/Documents/Project/Final_code")
```

# Install necessary packages

```{r}
# BiocManager::install("tximport")
# BiocManager::install("tximportData")
# BiocManager::install("DESeq2")
# install.packages('pheatmap')
# install.packages("DOSE")
# install.packages("enrichplot")
# install.packages("ggupset")
# BiocManager::install("biomaRt")
# BiocManager::install("pathview")
# BiocManager::install("EdgeR")
```

# Load packages

```{r}
library(biomaRt) # BiomaRt is used to access Ensembl databases for gene annotation
library(tximport) # Import transcript-level quantification data
library(readr) # For reading data files
library(DESeq2)
library(dplyr) # for data manipulation
library(PCAtools) # PCAtools is used for PCA analysis and visualization
library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
library(pheatmap)
library('org.Hs.eg.db') # Annotation package for human genes
library(GEOquery) # For downloading and processing data from GEO
library(edgeR) # For differential expression analysis
```

# Data import

## Import phenotype data

```{r}
metadata <- readxl::read_xlsx("../data/pbmc/pbmc_metadata.xlsx")
metadata <- as.data.frame(metadata)
metadata <- metadata[!is.na(metadata[[1]]),]

#Because RS10, RS15 kallisto files are missing, so we get rid of these samples in metadata
row.names(metadata) <- metadata$`RESOLVE Subject ID number`
metadata <-metadata[!row.names(metadata) %in% c("RS10","RS15"),]
head(metadata)
```

### Manipulate metadata to obtain necessary information

In this analysis, we use the information of estradiol level, BMI, age, so we use dplyr package to select and rename the columns we need

```{r}
metadata.subset <- metadata %>%
  dplyr::select(c(2,3,32,12)) %>%
  dplyr::rename(estradiol = 3,
         sex = 1,
         age = 2,
         BMI =4) %>%
  dplyr::mutate(estradiol = log(estradiol))

# check the data
head(metadata.subset)
```

```{r}
# filter for female samples only with estradiol level
metadata.subset <- metadata.subset %>%
  filter(sex == "Female")%>%
  filter(!is.na(estradiol))
```


## Import count data

```{r}
# Vector of folders of all samples
samples <- row.names(metadata.subset)
samples
sample_folders <- paste0(samples,"_PBMC_2")
sample_folders

# Create vector of paths to each abundance file
files <- file.path("../data/pbmc/pbmc_count_data/Kallisto abundance/PBMC",sample_folders,"abundance.tsv")
names(files) <- samples

#match transcripts and genes
tx2gene <- read.csv("../data/pbmc/pbmc_count_data/Kallisto abundance/gene_info.txt", sep="\t",header=T)
tx2gene <- tx2gene[,c(1,2)]
tx2gene

# load the Kallisto expression estimates using tximport
txi <- tximport(files,
                type = "kallisto",
                tx2gene = tx2gene,
                ignoreTxVersion = FALSE,
                countsFromAbundance = "lengthScaledTPM") 
```

# Differentially expressed genes analysis

## Create data object

```{r}
y <- DGEList(txi$counts, samples  = metadata.subset)
```


## Annotation
```{r}
#add gene symbol and gene name
library(org.Hs.eg.db)
library(AnnotationDbi)
all_gene_id <- row.names(y$counts)
annotation <- AnnotationDbi::mapIds(org.Hs.eg.db,
                     keys = all_gene_id,
                     column = "SYMBOL",
                     keytype = "ENSEMBL",
                     multiVals = "first") # multiVals = "first" will keep the first value for each gene if there are multiple values
```

```{r}
annotation <- data.frame(SYMBOL = annotation,
           stringsAsFactors = FALSE)
y$genes <- annotation
```


## Filtering and Normalization

Different Ensembl transcripts for the same gene symbol count predominantly the same reads.
So we keep one transcript for each gene symbol. We choose the transcript with highest
overall count:

```{r}
o <- order(rowSums(y$counts), decreasing=TRUE)
y <- y[o,]
```

```{r}
d <- duplicated(y$genes$SYMBOL)
y <- y[!d,]
nrow(y)
```

```{r}
# Calculate normalization factors
y <- calcNormFactors(y)
y
```

```{r}
# filtering using the design information
design <- model.matrix(~  estradiol + BMI + age , data = metadata.subset)
keep <- filterByExpr(y, design)
y <- y[keep, ]
```

## Differential expression

```{r}
vwts <- voom (y, design, plot=TRUE)
fit <- lmFit(vwts, design)
tmp <- contrasts.fit(fit, coef = 2)
tmp <- eBayes(tmp)
top.table <- topTable(tmp, coef = "estradiol", sort.by = "P", n = Inf)
head(top.table, 20)
```

```{r}
significant_genes <- sum(top.table$adj.P.Val < 0.05)
print(significant_genes)
```

So, with an adjusted p-value of 0.05, we do not identify any differentially expressed genes (DEGs) in the PBMC dataset.

```{r}
write.csv(top.table, file = "../result/pbmc/estrogen_continuous_DEGs_result.csv", row.names = TRUE) # save the results of all genes to a csv file
```


